Storm Data Analysis - Week 4

### 2021-03-24 Manuscript.Rmd
Completed event aggregation

# Make a copy of data and cast as DT
sel_data <- as.data.table(lean_data)
# Apply final transformations to DT
sel_data2 <- sel_data[, .(
    counter = .N, # Count elements in group
    # Total monetary damages per group
    total_dmg = sum(tot_dmg, na.rm = TRUE),
    # Total fatalities
    fatalities = sum(FATALITIES, na.rm = TRUE),
    # Injuries
    injuries = sum(INJURIES, na.rm = TRUE)
  ),
  # Grouping by event type
  by = .(event)][
    ,
    # Compute a "percent rank": rank events by number of
    # occurrences (desc), then assign a percentile by
    # dividing rank by denominator (i.e. nrows)
    # Note double dot syntax to access calling scope
    `:=`(perc_rank = frankv(counter)/.N,
         # Same for monetary damage
         dmg_rank = frankv(total_dmg)/..denominator,
         # And life damage
         inj_rank = frankv(injuries)/..denominator,
         fat_rank = frankv(fatalities)/..denominator
    )
  ][
    # Filter for frequency rank >=95th
    perc_rank >= 0.80
  ][
    order(-counter)
  ]

@@

Events below the 80^th^ percentile of frequency are mostly irrelevant to the analysis. To confirm this, we calculated total damages, injuries and fatalities from the global data and from the selected data. Results are identical (see above), despite the latter table having only a fraction of the event types (`r length(unique(gdata$event))` *vs.* `r length(unique(sel_data2$event))`).

@@

```{r event-aggregation}

@@

# Create regexp patterns to match to. The following is also the priority order
# with which events will be assigned to a category.
new_fcts <- list(
  hail = c("hail"),
  tornado = c("torn", "nado"),
  tsunami = c("tsun"),
  stormy = c("storm*", "thunder*", "tstm", "rain", "precip", "hurric", "spout"),
  lightning = c("lightn"),
  flood = c("flood", "seiche"),
  cold = c("cold", "ice", "sleet", "winter", "snow", "blizz", "avalan"),
  heat = c("heat", "hot", "summ", "warm"),
  windy = c("wind", "devil", "b*rst"),
  maritime = c("sea*", "ocean*", "mari[nt]*",
               "curr*", "tid*", "surf*", "wav", "swell"),
  fog = c("fog"),
  drought = c("drought"),
  fire = c("fire", "blaze", "smok"),
  landslide = c("slide")
)

@@

suppressWarnings(
  # For every new event category
  for (i in seq(1:length(new_fcts))) {
    # Match old event instances
    old_instances <- str_subset(sel_data$event, new_fcts[[i]])
    # Add to list
    old_events[[i]] <- unique(old_instances)
    i <- i + 1
  }
)
# Name (tag) list elements
names(old_events) <- names(new_fcts)

# TODO: switch to using list names instead of manual entry of new factor levels
sel_data$event2 <- fct_collapse(
  sel_data$event,
  hail = old_events$hail,
  tornado = old_events$tornado,
  tsunami = old_events$tsunami,
  stormy = old_events$stormy,
  lightning = old_events$lightning,
  flood = old_events$flood,
  cold = old_events$cold,
  heat = old_events$heat,
  windy = old_events$windy,
  maritime = old_events$maritime,
  fog = old_events$fog,
  drought = old_events$drought,
  fire = old_events$fire,
  landslide = old_events$landslide,
  other_level = "Other events"
) %>%
  fct_recode(
    "Hail" = "hail",
    "Winds" = "windy",
    "Tornadoes" = "tornado",
    "Floods" = "flood",
    "Lightnings" = "lightning",
    "Cold weather" = "cold",
    "Storms/Rains" = "stormy",
    "Oceanic events" = "maritime",
    "Fires" = "fire",
    "Droughts" = "drought",
    "Hot weather" = "heat",
    "Fog" = "fog",
    "Landslides" = "landslide",
    "Tsunamis" = "tsunami"
  ) %>%
  # Reorder according to frequency, desc
  fct_infreq()

@@

# Create plot data, use dtplyr for speed and convenience
pdata <- lazy_dt(sel_data)

>>>>

# Make a copy of data and cast as DT
sel_data <- as.data.table(lean_data)
# Apply final transformations to DT
sel_data2 <- sel_data[, .(
    counter = .N, # Count elements in group
    # Total monetary damages per group
    total_dmg = sum(tot_dmg, na.rm = TRUE),
    # Total fatalities
    fatalities = sum(FATALITIES, na.rm = TRUE),
    # Injuries
    injuries = sum(INJURIES, na.rm = TRUE)
  ),
  # Grouping by event type
  by = .(event)][
    ,
    # Compute a "percent rank": rank events by number of
    # occurrences (desc), then assign a percentile by
    # dividing rank by denominator (i.e. nrows)
    # Note double dot syntax to access calling scope
    `:=`(perc_rank = frankv(counter)/.N,
         # Same for monetary damage
         dmg_rank = frankv(total_dmg)/.N,
         # And life damage
         inj_rank = frankv(injuries)/.N,
         fat_rank = frankv(fatalities)/.N
    )
  ][
    # Filter for frequency rank >=95th
    perc_rank >= 0.80
  ][
    order(-counter)
  ]

@@

Events below the 80^th^ percentile of frequency are mostly irrelevant to the analysis. To confirm this, we calculated total damages, injuries and fatalities from the global data and from the selected data. The ratios of the selected data to the global data is near 1 (see above), despite the selected data table having only a fraction of the event types (`r length(unique(gdata$event))` *vs.* `r length(unique(sel_data2$event))`).

@@

```{r event-aggregation, warning=FALSE}

@@

# Create regexp patterns to match to. The following is also the priority order
# with which events will be assigned to a category.
new_fcts <- list(
  hail = c("hail"),
  tornado = c("torn", "nado"),
  tsunami = c("tsun"),
  stormy = c("storm*", "thunder*", "tstm", "rain", "precip", "hurric", "spout"),
  lightning = c("lightn"),
  flood = c("flood", "seiche", "fld"),
  cold = c("snow", "cold", "ice", "sleet", "^*inter", "blizz", "avalan"),
  heat = c("heat", "hot", "summ", "warm"),
  windy = c("wind", "devil", "b*rst"),
  fog = c("fog"),
  maritime = c("sea*", "ocean*", "mari[nt]*",
               "curr*", "tid*", "surf*", "wav", "swell"),
  drought = c("drought"),
  fire = c("fire", "blaze", "smok"),
  landslide = c("slide")
)

@@

suppressWarnings(
    # For every new event category
    for (i in seq(1:length(new_fcts))) {
        # Prepare storage array
        old_instances <- c()
        k <- 1
        # Loop over elements in list element
        for (k in seq(1:length(new_fcts[[i]]))) {
            # Match old event instances
            old_instances <- c(old_instances,
                               str_subset(sel_data2$event, new_fcts[[i]][k])
                                )
            k <- k + 1
        }
        # Add to list
        old_events[[i]] <- unique(old_instances)
        i <- i + 1
    }
)
# Name (tag) list elements
names(old_events) <- names(new_fcts)

# TODO: switch to using list names instead of manual entry of new factor levels
sel_data2$event2 <- fct_collapse(
  sel_data2$event,
  hail = old_events$hail,
  tornado = old_events$tornado,
  tsunami = old_events$tsunami,
  stormy = old_events$stormy,
  lightning = old_events$lightning,
  flood = old_events$flood,
  cold = old_events$cold,
  heat = old_events$heat,
  windy = old_events$windy,
  maritime = old_events$maritime,
  fog = old_events$fog,
  drought = old_events$drought,
  fire = old_events$fire,
  landslide = old_events$landslide,
  other_level = "Other events"
) %>%
  fct_recode(
    "Hail" = "hail",
    "Winds" = "windy",
    "Tornadoes" = "tornado",
    "Floods" = "flood",
    "Lightnings" = "lightning",
    "Cold weather" = "cold",
    "Storms/Rains" = "stormy",
    "Oceanic events" = "maritime",
    "Fires" = "fire",
    "Droughts" = "drought",
    "Hot weather" = "heat",
    "Fog" = "fog",
    "Landslides" = "landslide",
    "Tsunamis" = "tsunami"
  ) %>%
  # Reorder according to frequency, desc
  fct_infreq()

@@

# Create plot data, use dtplyr for speed and convenience
pdata <- lazy_dt(sel_data2)
